---
title: "CHIP Data Exploration"
execute:
  echo: false
author: Anthony Pompetti
date: "12-14-2023"
date-format: long
format: 
  revealjs:
    theme: simple
    logo: ../Coriell_Logo.png
    embed-resources: true
    view-distance: 20
---

```{r setup, include=FALSE}
library(data.table)
library(tidyverse)
library(plotly)
here::i_am(".here")
y <- readRDS(here::here("results/rds/CHIP_sample_table.rds"))
`%noin%` = Negate(`%in%`)
```

```{css cssFormat, echo=FALSE, include=FALSE}
.panel-tabset .nav-item {
  font-size: 3px
}
```

## Batches included in analysis
```{r}
y$batch %>% unique
```

## VAF by binned error {style="font-size: 16px;"}
#### minimal: (0 < age_delta <= 1), slight: (1 < age_delta <= 5), moderate: (5 < age_delta <= 10), high: (10 < age_delta <= 20), severe: (20 < age_delta) {style="font-size: 16px;"}
##### filter: (dp > 99 vaf > 0.02 impact != "MODIFIER" gene != "CBL") {style="font-size: 16px;"}

:::panel-tabset
### facet by error classifier
```{r, fig.width=20, fig.height=15, out.width="70%", fig.align = 'center'}
labels <- y %>% distinct(sample_id, .keep_all = T) %>% count(error_classifier)

labeller <- paste0(labels$error_classifier, ": n = ", labels$n)
names(labeller) <- labels$error_classifier
labeller <- as_labeller(labeller)

data <- y %>% 
  filter(
    impact %noin% c("MODIFIER") &
      gene != "CBL" &
      dp > 99 & 
      vaf > 0.02
  )

ggplot(
  data,
  aes(x = vaf,
      fill = impact)
) + 
  scale_fill_manual(
    values = c(HIGH = "red2", MODERATE = "purple2", LOW ="cyan2", MODIFIER = "khaki")
  ) +
  geom_histogram(aes(y=..density..), position = "stack") + 
  facet_wrap(~error_classifier, ncol = 2, labeller = labeller)
```

### facet by abserror classifier
```{r, fig.width=20, fig.height=15, out.width="70%", fig.align = 'center'}
labels <- y %>% distinct(sample_id, .keep_all = T) %>% count(abserror_classifier)

labeller <- paste0(labels$abserror_classifier, ": n = ", labels$n)
names(labeller) <- labels$abserror_classifier
labeller <- as_labeller(labeller)

data <- y %>% 
  filter(
    impact %noin% c("MODIFIER") &
      gene != "CBL" &
      dp > 99 & 
      vaf > 0.02
  )

ggplot(
  data,
  aes(x = vaf,
      fill = impact)
) + 
  scale_fill_manual(
    values = c(HIGH = "red2", MODERATE = "purple2", LOW ="cyan2", MODIFIER = "khaki")
  ) +
  geom_histogram(aes(y=..density..), position = "stack") + 
  facet_wrap(~abserror_classifier, ncol = 2, labeller = labeller)
```
:::

## Error of variants by impact {style="font-size: 16px;"}
##### filter: (dp > 99 vaf > 0.02 impact != "MODIFIER" gene != "CBL") {style="font-size: 16px;"}
```{r, fig.width=10, fig.height=8, fig.align = 'center'}
data <- y %>% 
  filter(impact %noin% c("MODIFIER") & gene != "CBL" & dp > 99 & vaf > 0.02)

p <- ggplot(
  data, 
  aes(x = impact, y = error, fill = impact),
) +
  geom_boxplot(alpha = .6) +
  scale_fill_manual(
    values = c(HIGH = "red2", MODERATE = "purple2", LOW ="cyan2", MODIFIER = "khaki")
  )

p
```

## Abserror of variants by impact {style="font-size: 16px;"}
##### filter: (dp > 99 vaf > 0.02 impact != "MODIFIER" gene != "CBL") {style="font-size: 16px;"}
```{r, fig.width=10, fig.height=8, fig.align = 'center'}
data <- y %>% 
  filter(impact %noin% c("MODIFIER") & gene != "CBL" & dp > 99 & vaf > 0.02)

p <- ggplot(
  data, 
  aes(x = impact, y = abserror, fill = impact),
) +
  geom_boxplot(alpha = .6) +
  scale_fill_manual(
    values = c(HIGH = "red2", MODERATE = "purple2", LOW ="cyan2", MODIFIER = "khaki")
  )

p
```

## Chip relevant variants per sample {style="font-size: 16px;"}
#### filter: (dp > 99 & vaf > 0.02 & vaf < 0.4 impact != "MODIFIER" gene != "CBL") {style="font-size: 16px;"}
::: panel-tabset
### scatterplot by error
```{r, fig.align = 'center'}
data <- y %>% 
  filter(impact %noin% c("MODIFIER") & gene != "CBL") %>%
  dplyr::group_by(sample_id, chip_relevant, .drop = FALSE) %>% 
  count() %>%
  dplyr::left_join(., y %>% 
                     select(sample_id, abserror, error, batch, abserror_classifier, error_classifier), 
                   by = "sample_id") %>%
  select(sample_id, chip_relevant, n, abserror, error, batch, abserror_classifier, error_classifier) %>%
  distinct(sample_id, chip_relevant, .keep_all = TRUE) %>%
  filter(chip_relevant == "TRUE")

p <- ggplot(
  data, 
  aes(x = error, y = n, color = batch, custom = sample_id)
) +
  geom_point()

p
```

### scatter plot by abserror
```{r, fig.align = 'center'}
data <- y %>% 
  filter(impact %noin% c("MODIFIER") & gene != "CBL") %>%
  dplyr::group_by(sample_id, chip_relevant, .drop = FALSE) %>% 
  count() %>%
  dplyr::left_join(., y %>% 
                     select(sample_id, abserror, batch, abserror_classifier, error_classifier), 
                   by = "sample_id") %>%
  select(sample_id, chip_relevant, n, abserror, batch, abserror_classifier, error_classifier) %>%
  distinct(sample_id, chip_relevant, .keep_all = TRUE) %>%
  filter(chip_relevant == "TRUE")

p <- ggplot(
  data, 
  aes(x = abserror, y = n, color = batch, custom = sample_id)
) +
  geom_point()

p
```
:::

## Chip relevant variants per sample binned by error {style="font-size: 16px;"}
#### filter: (dp > 99 & vaf > 0.02 & vaf < 0.4 impact != "MODIFIER" gene != "CBL") {style="font-size: 16px;"}
::: panel-tabset
### boxplot binned by error_classifier
```{r, fig.width=15, fig.align = 'center'}
data <- y %>% 
  dplyr::group_by(sample_id, chip_relevant, .drop = FALSE) %>% 
  count() %>%
  dplyr::left_join(., y %>% 
                     select(sample_id, abserror, batch, abserror_classifier, error_classifier), 
                   by = "sample_id") %>%
  select(sample_id, chip_relevant, n, abserror, batch, abserror_classifier, error_classifier) %>%
  distinct(sample_id, chip_relevant, .keep_all = TRUE) %>%
  filter(chip_relevant == "TRUE")


p <- ggplot(
  data, 
  aes(x = factor(error_classifier, levels = levels(y$error_classifier)), y = n)
) +
  geom_boxplot() +
  xlab("error_classifier")

p
```

### boxplot binned by abserror_classifier
```{r, fig.align = 'center'}
data <- y %>% 
  dplyr::group_by(sample_id, chip_relevant, .drop = FALSE) %>% 
  count() %>%
  dplyr::left_join(., y %>% 
                     select(sample_id, abserror, batch, abserror_classifier, error_classifier), 
                   by = "sample_id") %>%
  select(sample_id, chip_relevant, n, abserror, batch, abserror_classifier, error_classifier) %>%
  distinct(sample_id, chip_relevant, .keep_all = TRUE) %>%
  filter(chip_relevant == "TRUE")



p <- ggplot(
  data, 
  aes(x = factor(abserror_classifier, levels = levels(y$abserror_classifier)), y = n)
) +
  geom_boxplot() +
  xlab("abserror_classifier")

p
```
:::
